---
title: "Dimension Reduction</br>Isomap"
author: "Anastasios Panagiotelis"
institute: "University of Sydney"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css: [default,"../css/mtheme.css","../css/mod.css"]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    includes:
      before_body: ../auxil/defs.html
---

#Outline

- Geodesics
--

  + Approximation using NN graph
--

- Isomap algorithm
--

- Applications
--

- Initially proposed in J. B. Tenenbaum, V. de Silva, J. C. Langford, (2000), "A Global Geometric Framework for Nonlinear Dimensionality Reduction", *Science* **290**, pp. 2319–2323.

```{r setup, include=FALSE,message=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.align='center',echo=FALSE,message = FALSE,out.height = 450)

library(gifski)
library(knitr)
library(dimRed)
library(rgl)
knit_hooks$set(webgl = hook_webgl)

```

---

class: center, middle, inverse

# Isomap

---

# Idea

- Recall MDS takes distances as inputs.
--

- These are typically Euclidean
--

- If non-Euclidean distances are used as inputs, then the output still represents these distances quite accurately.
--

- ISOMAP uses geodesic distances along the manifold as inputs.

---

# Toy example

Consider the following data

```{r}
library(tidyverse)
library(plotly)
pts<-400
t<-4*pi*(1-exp(-(seq(0,2,length.out = pts))))
x<-rep(NA,pts)
y<-rep(NA,pts)
for (tt in 1:pts){
  x[tt]<-t[tt]*cos(t[tt])
  y[tt]<-t[tt]*sin(t[tt])
}
df<-tibble(t,x,y)
df_sub<-filter(df,row_number()%%10==1)

df_sub%>%
  ggplot(aes(x=x,y=y))+geom_point(size=4)+theme_void()

```

---


# Manifold

Lying on a manifold.

```{r}
df%>%
  ggplot(aes(x=x,y=y))+geom_path(color="#E69F00",size=2)+geom_point(data=df_sub,aes(x=x,y=y),size=4)+theme_void()

```


---



# Euclidean distance

Blue points close in ambient space.

```{r}
df_red<-slice(df_sub,c(9,38))
df%>%
  ggplot(aes(x=x,y=y))+geom_path(color="#E69F00",size=2)+
  geom_point(data=df_sub,aes(x=x,y=y),size=4)+
  geom_point(data=df_red,aes(x=x,y=y),size=4,col="#56B4E9")+
  geom_line(data=df_red,aes(x=x,y=y),size=2,col="#56B4E9")+
  theme_void()

```



---

# Geodesic Distance

But not on manifold.

```{r}
df_ps<-slice(df,c(81:370))

df%>%
  ggplot(aes(x=x,y=y))+geom_path(color="#E69F00",size=2)+
  geom_point(data=df_sub,aes(x=x,y=y),size=4)+
  geom_point(data=df_red,aes(x=x,y=y),size=4,col="#56B4E9")+
  geom_line(data=df_red,aes(x=x,y=y),size=2,col="#56B4E9")+
  geom_path(data=df_ps,aes(x=x,y=y),color= "#009E73",size=2)+
  theme_void()

```

---

# Input distances

- Classical MDS would use the distance in blue as an input
--

- The idea behind Isomap is to use the distance in green as an input.
--

- However to compute the geodesic (green) distance we need to know the manifold.
--

- Instead we approximate the geodesic distance

---

# Geodesic Distance

Geodesic can be approximated.

```{r}
df_da<-slice(df_sub,c(9:38))

df%>%
  ggplot(aes(x=x,y=y))+geom_path(color="#E69F00",size=2)+
  geom_point(data=df_sub,aes(x=x,y=y),size=4)+
  geom_path(data=df_da,aes(x=x,y=y),size=1)+
  geom_point(data=df_red,aes(x=x,y=y),size=4,col="#56B4E9")+
  geom_line(data=df_red,aes(x=x,y=y),size=2,col="#56B4E9")+
  geom_path(data=df_ps,aes(x=x,y=y),color= "#009E73",size=2)+
  geom_path(data=df_da,aes(x=x,y=y),size=1)+
  theme_void()

```

---


# Geodesic Distance

Try zooming.

```{r}
library(widgetframe)
  
df_da<-slice(df_sub,c(9:38))

df%>%
  ggplot(aes(x=x,y=y))+geom_path(color="#E69F00",size=1)+
  geom_point(data=df_sub,aes(x=x,y=y),size=4)+
  geom_point(data=df_red,aes(x=x,y=y),size=4,col="#56B4E9")+
  geom_line(data=df_red,aes(x=x,y=y),size=1,col="#56B4E9")+
  geom_path(data=df_ps,aes(x=x,y=y),color= "#009E73",size=2)+
  geom_path(data=df_da,aes(x=x,y=y),size=1)+
  theme_void()->g

ggplotly(g)%>%
  frameWidget()

```

---

class: center, middle, inverse

# Algorithm

---

# Step 1: Nearest Neighbours

- Find the neighbourhood graph
--

  + Neighbourhood within a ball $\epsilon$-Isomap
  + Neearest neighbours $K$-Isomap
--

- As edge weights use the Euclidean distance between points
  

---


# Example


```{r,message=F,echo=F}
library(RANN)
library(igraph)
library(ggrepel)
set.seed(5)
h<-rep('reg',10)
h[3]<-'point'
h[c(2,6,9)]<-'nn'
df<-tibble(X1=runif(10),X2=1-runif(10))
df%>%ggplot(aes(x=X1,y=X2,label=as.character(1:10)))+geom_point(size=6)+geom_text_repel(size=8)+coord_equal()

```

---

# Graph

```{r,echo=F}
nn<-nn2(df)
k<-3
dfe<-tibble(N1=rep(1:10,k),N2=as.vector(nn$nn.idx[,2:(k+1)]))
graph_from_data_frame(dfe,directed = FALSE)%>%
  simplify(remove.multiple = TRUE)->ng
dd<-dist(cbind(df$X1,df$X2))%>%as.matrix()
wt<-rep(NA,length(E(ng)))
for(i in 1:length(E(ng))){
  nds<-ends(ng,i)%>%as.numeric()
  wt[i]<-dd[nds[1],nds[2]]
  
}
edge_attr(ng,'weight')<-wt
plot(ng,edge.width=5*wt)
```

---

# Step 2: Shortest Path

- Find shortest path on graph, between every pair of points.
--

- Algorithms to do this
--

  + Dijkstra's method
  + Floyd- Warshall algorithm
  + and more...
--

- Let's look at a simple example for one pair

---

# Path from 8 to 10

```{r}
plot(ng,edge.width=5*wt)
```
---
# Shortest path

```{r}
plot(ng,edge.width=5*wt,mark.groups=list(c(10,4),c(4,7),c(7,8)),mark.col="#CCCCCC",mark.border="#000000")
```
---

# On original plot

```{r,message=FALSE}

ggplot(df,aes(x=X1,y=X2))+
  geom_point(size=6)+
  geom_text_repel(size=8,label=as.character(1:10))+
  coord_equal()

```
---

# On original plot

```{r,message=FALSE}
df_sub<-slice(df,c(10,4,7,8))

ggplot(df,aes(x=X1,y=X2))+
  geom_point(size=6)+
  geom_text_repel(size=8,label=as.character(1:10))+
  geom_path(data=df_sub,mapping = aes(x=X1,y=X2),col='blue',size=2)+
  coord_equal()

```

---

# Approximation

- The approximate geodesic distance is given by the sum of the edge weights along the shortest path.
--

- The quality of this approximation depends on
  + Size of $\epsilon$ or $K$
  + The density of points $\alpha$
  + Curvature of manifold
  + Dimension of ambient space
  + Connectedness of graph

---

# More precisely

For arbitrarily small $\lambda_1$, $\lambda_2$ and $\mu$, with probability $1-\mu$ 

$$\delta_\calM(\bx_i,\bx_j)(1-\lambda_1)\leq\hat{\delta}_\calM(\bx_i,\bx_j)\\\leq\delta_\calM(\bx_i,\bx_j)(1+\lambda_2)$$
if some conditions hold for $\epsilon$ (or $K$) and the density $\alpha$ of points on the manifold. 

---

# Step 3: MDS

- Carry out classical MDS however...
--

- Use the approximate geodesic distances as inputs
--

- Notes on MDS for more details.

---

class: center, middle, inverse

# Application

---

# S Curve

```{r,echo=T,eval=F}
dat <- loadDataSet("3D S Curve")
plot(dat,type='3varsrgl')

```

```{r,echo=F,webgl=TRUE}

dat <- loadDataSet("3D S Curve")
invisible(open3d())
plot(dat,type='3varsrgl')
rglwidget()
close3d()


```

---

# MDS

```{r,echo=T}
mdsout <- embed(dat, "MDS")
plot(mdsout, type = "2vars")

```

---

# Isomap (K=20)

```{r,echo=T}
isoout <- embed(dat, "Isomap", knn = 20)
plot(isoout, type = "2vars")

```

---

# Isomap (K=4)

```{r,echo=T}
isoout <- embed(dat, "Isomap", knn = 4)
plot(isoout, type = "2vars")

```

---

# Isomap (K=500)

```{r,echo=T,cache=T}
isoout <- embed(dat, "Isomap", knn = 500)
plot(isoout, type = "2vars")

```

---

# Pros and Cons

- Advantages
  + Exploits geometry
  + Global in the sense that all pairwise geodesic distances are used.
--

- Disadvantages
  + Floyd Warshall algorithm is $O(n^3)$
  + Eigendecomposition for MDS is $O(n^3)$
  + Not clear how to pick $K$ or $\epsilon$ 


---

class: center, middle, inverse

#Questions?