---
title: "Dimension Reduction</br>Determining Dimension"
author: "Anastasios Panagiotelis"
institute: "University of Sydney"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css: [default,"../css/mtheme.css","../css/mod.css"]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    includes:
      before_body: ../auxil/defs.html
---

#Outline

- Spheres
--

- Applications
--

- Reference is Facco, E., d’Errico, M., Rodriguez, A., & Laio, A. (2017). Estimating the intrinsic dimension of datasets by a minimal neighborhood information. *Scientific reports*, **7(1)**, 1-8.

```{r setup, include=FALSE,message=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.align='center',echo=FALSE,message = FALSE,out.height = 450)

library(gifski)
library(tidyverse)
library(knitr)
library(ggthemes)
library(dimRed)
library(rgl)
library(ggrepel)
library(plotrix)
library(microbenchmark)
knit_hooks$set(webgl = hook_webgl)

```

---

class: center, middle, inverse

# What is dimension

---

# Lengths and volumes

- Consider a 1-D 'ball'.
--

- This is just a line.
--

- What is the measure of a 1-D ball?
--

  + The measure is length and is $2r$ where $r$ is the radius.
--

- What happens the the measure if we double radius of the 1-D ball? 

---


# In 2-D

- Consider a 2-D 'ball'.
--

- What is the measure of a 2-D ball?
--

  + The measure is area and is $\pi r^2$ where $r$ is the radius.
--

- What happens the the measure if we double the radius of the 2-D ball? 
--

  + Measure scaled up by $2^2=4$.

---


# In 3-D

- Consider a 3-D 'ball'.
--

- What is the measure of a 3-D ball?
--

  + The measure is volume and is $\frac{4}{3}\pi r^3$ where $r$ is the radius.
--

- What happens the the measure if we double the radius of the 3-D ball? 
--

  + Measure scaled up by $2^3=8$.

---

# Why does this matter?

- Consider the nearest neighbour and second nearest neighbour of a point.
--

- Consider a ball around the second nearest neighbor
```{r,fig.height=6}
plot(0,0,col='red',ylim=c(-1,1),pch=20,cex=3,xlim=c(-1,1),ylab='',xlab='',asp=1)
points(0.6,0.8,col='blue',pch=20,cex=3)
points(0.3,-0.4,col='black',pch=20,cex=3)
draw.circle(0,0,1)
```
---


# Why does this matter?

- Consider the nearest neighbour and second nearest neighbour of a point.
- Consider a ball around the second nearest neighbor
```{r,fig.height=6}
plot(0,0,col='red',ylim=c(-1,1),pch=20,cex=3,xlim=c(-1,1),ylab='',xlab='',asp=1)
points(0.6,0.8,col='blue',pch=20,cex=3)
points(0.3,-0.4,col='black',pch=20,cex=3)
lines(c(0,0.6),c(0,0.8),col='blue')
lines(c(0,0.3),c(0,-0.4))
draw.circle(0,0,1)
draw.circle(0,0,0.5,lty = 3)
```

---

# Questions

- In the previous example, the ratio of the distance to the second nearest neighbour to distance to the nearest neighbour is 2.
--

- What is the probability that this ratio is greater than 2?
--

- To answer calculate area of inner circle and divide by outer circle.
--

- It is 0.25.
--

- How would your answer change if it was a 3D ball?

---

# Determine dimension

- For each point find the ratio of the second nearest neighbour to the nearest neighbour (denoted $\mu$)
- These follow a Pareto(1,d) distribution. The MLE is given by

$$\hat{d}=\frac{n-1}{\sum\limits_{i=1}^{n} log \mu_i}$$
---

# Code

```{r,echo=T}
data<-loadDataSet("Swiss Roll")
dd<-as.matrix(dist(data@data))
dr<-apply(dd,2,sort)
mu<-dr[3,]/dr[2,]
dhat<-(length(mu)-1)/sum(log(mu))
dhat
```

---

#Helix

```{r,webgl=T}
t<-runif(1000)*20
x<-cos(t)
y<-sin(t)
z<-t
plot3d(x,y,z)
```


---

# Helix

```{r,echo=T}
data<-cbind(x,y,z)
dd<-as.matrix(dist(data))
dr<-apply(dd,2,sort)
mu<-dr[3,]/dr[2,]
dhat<-(length(mu)-1)/sum(log(mu))
dhat
```

---

#Noisy Helix

```{r, webgl=T}
data<-loadDataSet("variable Noise Helix")
plot(data,type='3varsrgl')
```

---

# Code

```{r,echo=T}
dd<-as.matrix(dist(data@data))
dr<-apply(dd,2,sort)
mu<-dr[3,]/dr[2,]
dhat<-(length(mu)-1)/sum(log(mu))
dhat
```

---

# Conclusions

- The intrinsic dimension can be estimated using nothing more than nearest neighbour information.
--

- Can be generalised to include more than 2 NN
--

- Good idea to trim large values of the ratio of NN.
--

- Method is not robust if too much noise is added.
